Towards a Complete Model of the Negative Selection Process in Humans

### 2023-05-22 manuscript/manuscript.tex
ms changes

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \parencite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent twice the rate of the ratchet of deleterious mutations in fixations
per generation, each fixation removes exactly the variation mutation puts in
$V_m = Us$ (we omit the segment index for clarity). Thus, under a general
equilibrium negative selection process,

\begin{align}
  \label{eq:Va}
  V_{a} = (U - R)s. 
\end{align}

Note that since mutation is the ultimate source of all alleles that are fixed
by the ratchet or are lost, $U \ge R$. Intuitively, this equilibrium genic
fitness variation is a balance between the rate of gain of genic fitness
variation due to mutation and its loss due to fixation. Our results differ from
\parencite{Santiago2016-mu}, since we find from simulations their expression
models the additive \emph{genic} fitness variation, not additive genetic
fitness variation (see Section XXX), since it does not adequately capture
linkage disequilibria dynamics. Santiago and Caballero demonstrate Equation
\eqref{eq:Va} through Fisher's Fundamental Theorem of Natural Selection, but we
find an alternative proof (\cite{Higgs1995-xc}; see Supplementary Materials
Section \ref{supp:weak-strong}).

However, closed-form expressions for the rate of the ratchet $R$ have generally
been hard to find \parencite{Haigh1978-gt,Higgs1995-xc,Gessler1995-hz}. The key
insight of \parencite{Santiago2016-mu} is that the ratchet rate under draft is
given by the probability of fixation $p_f(N_d, s)$
\parencite{Kimura1962-su,Malecot1948-zv} with the draft-rescaled effective
population size, e.g. $R = N U p_f(N_d, s)$. Given this equation for the
ratchet and Equation \eqref{eq:polygenic_Ne} for $N_d$ under draft, we have a
system of two non-linear equations that can be solved numerically for $N_d$ and
$R$ for each segment,

@@

Given that Santiago and Caballero's (\citeyear{Santiago2016-mu}) model is a
haploid-model for a single region under equilibrium negative selection, we
first sought to confirm it could accurately predict the reduction in realistic
forward-time simulations of diploid populations with recombination. Using SLiM
\parencite{Haller2019-vu,Haller2023-uk} we simulated a region of $10^5$
basepairs under varying levels of mutation and selection. We find a close
correspondence between the observed and predicted reduction in effective
population size $B=\nicefrac{N_d}{N}$ over all selection and mutation
parameters including weak selection (Figure 1A), in contrast to classic BGS
theory. Furthermore, to investigate whether this accuracy was caused by the
model capturing the equilibrium fitness variance and ratchet rate, we also
measured these throughout the simulation. Again, we find their theory
accurately predicts both the ratchet rate (Figure 1B) and the genic fitness
variance (Figure 1C). Additionally, these figures provide further intuition for
the underlying negative selection process. When selection is strong the ratchet
rate is zero, as there is no probability that strongly deleterious can fix.
However, around $2N_e s \approx 1$, the ratchet begins clicking as $p_F > 0$.
Ultimately, the ratchet rate reaches the neutral mutation rate, as $V_a = 0$
and $N_d = N$ (e.g. no draft). Likewise, the additive genic fitness variance
matches the level predicted by BGS ($Us$), but begins to diverge when the
ratchet begins clicking.

@@

Given the accuracy of the SC16 model in predicting the reduction in effective
population size and ratchet rate when equilibrium negative selection processes
were acting in single segment, we extended their model so that it could be
applied to whole-genome data.

@@

\begin{align}
  V_{A} = Us - Rs. 
\end{align}

where $U = 2u$ and $R=2r$ is the diploid rate of the ratchet, which depends on
higher moments of the fitness distribution, but only differs from $r$ through a
rescaling of the population size. Note that if selected sites are sufficiently
deleterious that they cannot fix (or the population is infinite), $R=0$ and the
additive fitness variance is identical to that under strong background
selection, $V_A = Us = V_\text{BGS}$ (equation \eqref{eq:va_bgs}). Then, the
fixation probability of a new deleterious mutation with additive effects ($1-s$
in the heterozygote, $1-2s$ in the homozygote) in a diploid population is,

@@

>>>>

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \parencite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent the rate of the ratchet of deleterious mutations in fixations per
generation, each fixation removes the equivalent amount of equilibrium
variation put in by mutation. Thus, the steady-state genic variance under
mutation and negative selection is (omitting the segment index $i$ for
clarity),

\begin{align}
  \label{eq:Va}
  V_{a} = (U - 2 R)s. 
\end{align}

where the condition $V_a \ge 0$ is met when the probability of fixation is less
than or equal to the neutral fixation probability of $\nicefrac{1}{2N}$, which
is held when considering deleterious mutations. This equation connects the
equilibrium additive genic fitness variance to the flux of new variation in due
to new deleterious mutations and the flux out due to their substitution and the
decline in mean population fitness. When $R=0$, the selection is so strong it
cannot fix, and the equilibrium fitness variation is due entirely to young rare
mutations before their extinction $V_a = V_a^{BGS} = Us$. Santiago and
Caballero demonstrate Equation \eqref{eq:Va} through Fisher's Fundamental
Theorem of Natural Selection, but we find an alternative proof
(\cite{Higgs1995-xc}; see Supplementary Materials Section
\ref{supp:weak-strong}). We find that the steady-state additive genic variance
in Equation \eqref{eq:Va} also results from diffusion models with a flux of
mutations into discrete sites (\cite{Kimura1969-jw}; see Supplementary
Materials Section \ref{supp:kimura}). Our results differ from
\parencite{Santiago2016-mu}, since we find from simulations their expression
models the additive \emph{genic} fitness variation, not additive genetic
fitness variation (see Section XXX). This is because Equation \eqref{eq:Va}
does not capture the expected build-up of negative linkage disequilibria when
selection is only mildly deleterious.

While using Equation \eqref{eq:Va} in Equation \eqref{eq:main_Ne} leads to a
prediction for the draft-effective population size $N_d$, closed-form
expressions for the rate of the ratchet $R$ have generally been hard to find
\parencite{Haigh1978-gt,Higgs1995-xc,Gessler1995-hz}. The key insight of
\parencite{Santiago2016-mu} is that the ratchet rate under draft is given by
the probability of fixation $p_f(N_d, s)$
\parencite{Kimura1962-su,Malecot1948-zv} with the draft-rescaled effective
population size, e.g. $R = N U p_f(N_d, s)$. Given this equation for the
ratchet and Equation \eqref{eq:polygenic_Ne} for $N_d$ under draft, we have a
system of two non-linear equations that can be solved numerically for $N_d$ and
$R$ for each segment,

@@

Given that Santiago and Caballero's (\citeyear{Santiago2016-mu}) model is a
haploid-model for a single region under equilibrium negative selection, we
first sought to confirm it could accurately predict the reduction in realistic
forward-time simulations of diploid populations with recombination. Using SLiM
\parencite{Haller2019-vu,Haller2023-uk} we simulated a region of $10^5$
basepairs under varying levels of mutation and selection. We find a close
correspondence between the observed and predicted reduction in effective
population size $B=\nicefrac{N_d}{N}$ over all selection and mutation
parameters including weak selection (Figure 1A), in contrast to classic BGS
theory. Furthermore, to investigate whether this accuracy was caused by the
model capturing the equilibrium fitness variance and ratchet rate, we also
measured these throughout the simulation. Again, we find their theory
accurately predicts both the ratchet rate (Figure 1B) and the genic fitness
variance (Figure 1C).

Additionally, these figures provide further intuition for the underlying
negative selection process. When selection is strong the ratchet rate is zero,
as there is no probability that strongly deleterious can fix. However, around
$2N_e s \approx 1$, the ratchet begins clicking as $p_F > 0$. Ultimately, the
ratchet rate reaches the neutral mutation rate, as $V_a = 0$ and $N_d = N$
(e.g. no draft). Likewise, the additive genic fitness variance matches the
level predicted by BGS ($Us$), but begins to diverge when the ratchet begins
clicking. This occurs when the level of variation diverges from the
deterministic mutation-selection equilibrium $V_a = Us$ due to the stochastic
clicking of the ratchet at rate $R$.

@@

Given the accuracy of the SC16 model in predicting the reduction in effective
population size and ratchet rate when equilibrium negative selection processes
were acting in single segment, we extended their model so that it could be
applied to whole-genome data. Our method, implemented in the software
\texttt{bgspy}, calculates the equilibrium additive genic fitness variance
$V_a$ and ratchet rate $R$ for each specified segment in the genome for a range
of mutation rates and selection coefficients and a fixed genome-wide
drift-effective population size $N$ (see Methods XXX). Genomic regions
putatively under negative selection (e.g. coding sequences or UTRs) and the
recombination rate map are specified as arguments so that the predicted values
accurately reflect local genomic details. Note that each equation is solved
independently, since here we assume no selective interference between segments.

To assess our genome-wide extension of the SC16 model, we simulated negative
selection for fixed mutation rates and selection coefficient on human
chromosome 10, using the HapMap recombination map (XXX) and in coding regions,
PhastCons regions, and UTRs as an approximate set of putatively conserved
segments. We then compared the reduction over an average of XXX simulation
replicates with the predicted reduction using our method.

We find that our theory closely matches simulations when selection is strong
(Figure 1XXX and Figure 1XXX). Since our theory is identical to classic BGS in
this domain, this provides the first genome-wide realistic forward-time
simulation confirmation of this theory. However, we find slight discrepancies
in regions with low recombination (Figure XXX). Second, we find our theory is
vastly more accurate than classic BGS theory when selection is weak ($2N_e s
\ll 1$; Figure 1XXX). However, when the strength of selection is at the
drift-barrier of $2N_e s \approx 1$), we find our theory has about a 10\%
relative error rate, which is slightly lower than classic BGS theory.

We first hypothesized that this discrepancy could be because slight selective
interference between segments

could be due to the build up of negative
linkage disequilibrium

@@

\begin{align}
  \label{supp-eq:Va}
  V_{A} = Us - 2Rs. 
\end{align}

where $U = 2u$ and $R=r$ is the diploid rate of the ratchet, which depends on
higher moments of the fitness distribution, but only differs from $r$ through a
rescaling of the population size. Note that if selected sites are sufficiently
deleterious that they cannot fix (or the population is infinite), $R=0$ and the
additive fitness variance is identical to that under strong background
selection, $V_A = Us = V_\text{BGS}$ (equation \eqref{eq:va_bgs}). Then, the
fixation probability of a new deleterious mutation with additive effects ($1-s$
in the heterozygote, $1-2s$ in the homozygote) in a diploid population is,

@@

\subsection*{Steady-state Equilibrium Genic Variance} \label{supp:genic-va}

We also find that Equation \eqref{eq:Va} (and Supplementary Materials Equation
\ref{supp-eq:Va} above) can also be found through Kimura's diffusion models
with a flux of mutations into infinite, discrete sites
(\citeyear{Kimura1969-jw}). This model integrates the Kolmogorov backwards
equation over time and forward variable to find the expected value of a
functional of the initial frequency. Using the functional $I(p) = s^2 p(1-p)$
for genic variance given mutations start at frequency $p$, one finds that the
steady-state genic variance for $p=\nicefrac{1}{2N_e}$ is

\begin{align}
    V_a^{SS} &= 2 U N s \left(\frac{1}{2N_e} - p_f(N_e, s)\right) \\
             &= \left(\frac{UN}{N_e} - 2 U Np_f(N_e, s))\right)s \nonumber \\
             &\approx \left(U - 2 R\right)s \nonumber
\end{align}

which is approximately Equation \eqref{eq:Va}.
